---
title: "Basic command and data types"
author: "Hassan Ghayas"
date: '2026-04-26'
format:
  html:
    self-contained: true
    toc: true            # Enables the Table of Contents
    toc-location: left   # Moves the TOC to the left (default is right)
    toc-depth: 3         # Sets how many levels of headings to show
editor: visual
---

## **Variable**

Variables are objects in R use to store values. It can be single value, basic or complex arithmetic operations, or a data matrix or a data frame. Variable can be assign using either the "**\<-**" or "**=**" operators. (alt+ -)

```{r}
a = 5
b <- 8*3
c = "Data analysis with R"
a + b
c
```

```{r}
cat(a,c)
```

## Basic command

```{r}
x <-  3
y <-  2
```

```{r}
x+y
```

```{r}
class(x)
class(y)
x <- as.integer(x)
class(x)
```

```{r}
sqrt(3 * 4 + 2 * 5 + 3)
```

```{r}
log(5) + log(10)
```

```{r}
log(50) 
```

```{r}
sum(3 * 4 + 2 * 5 + 3)
```

## Data Type

Variables can contain values of specific types within R. The six **data types** that R uses include:

-   `"numeric"` ---\> any numerical value, including whole numbers and decimals.

-   `"character"` ---\> text values, denoted by using quotes ("") around text.

-   `"integer"` ---\> whole numbers (e.g., `2L`, the `L` indicates to R that it's an integer). It behaves similar to the `numeric` data type for most tasks or functions; however, it takes up less storage space than numeric data.

-   `"logical"` ---\> `TRUE` and `FALSE` (the Boolean data type). The `logical` data type can be specified using four values, `TRUE` in all capital letters, `FALSE` in all capital letters, a single capital `T` or a single capital `F`.

## Data Structure

### Vector

A vector is substantially a list of variables. A vector consists of a collection of numbers, arithmetic expressions, logical values or character strings for example. However, each vector must have all components of the same type, i.e numeric, character, or logical

```{r}
c(3,8,2)
c("mike", "John")
c(5,9,"abc", TRUE)
```

```{r}
numbers <- c(38,26,35)
name <- c("mike", "john", "jane")
any_text <- c(5,9,"abc", TRUE)
```

```{r}
class(numbers)
class(name)
class(any_text)
```

### Data Frame

Two-dimensional, tabular data structure used to store data where each column represents a variable and each row represents an observation

```{r}
df <- data.frame(name, numbers)
```

```{r}
head(df)
```

```{r}
colnames(df)[2] <- "age"
```

```{r}
names(df)[names(df)=="numbers"] <- "age"
```

```{r}
summary(df$age)
```

### Matrix

A **matrix** in R is a collection of vectors of **same length and identical datatype**

```{r}
data_matrix <- as.matrix(cbind(a=numbers,b=2*numbers,c=numbers/1.5))
```

```{r}
data_matrix
```
