---
title: "Basic Statistical Analysis"
author: "Hassan Ghayas"
date: '2026-04-27'
format:
  html:
    self-contained: true
    toc: true            # Enables the Table of Contents
    toc-location: left   # Moves the TOC to the left (default is right)
    toc-depth: 4         # Sets how many levels of headings to show
editor: visual
---

## 📦 Load Libraries

```{r}
library(dplyr)
library(ggplot2)
library(palmerpenguins)

```

### Load data

For basic statistical analysis, we will use palmerpenguins r package which has penguin dataset

```{r}
penguin_data <- penguins
head(penguin_data)
```

### Summary Statistics

Descriptive statistics are used to summarize and describe the main features of a dataset. Common descriptive statistics include:

-   Mean

-    Median

-    Standard deviation

-    Minimum and maximum values

```{r}
mean(penguin_data$body_mass_g, na.rm = TRUE)
```

```{r}
median(penguin_data$body_mass_g, na.rm = TRUE)
```

```{r}
sd(penguin_data$body_mass_g, na.rm = TRUE)
```

```{r}
summary(penguin_data$body_mass_g)
```

```{r}
penguin_data %>%
  group_by(species) %>%
  summarise(mean_mass = mean(body_mass_g, na.rm = TRUE))
```

### t-test

A t-test is used to compare the means between two groups to determine whether they are statistically different.

Required:

-   One numeric variable

-   One grouping variable with two categories

```{r}
t.test(body_mass_g ~ sex, data = penguin_data)
```

### ANOVA

ANOVA is used to compare the means among three or more groups. For example comparing body mass among species

Required:

-   One numeric variable

<!-- -->

-    One categorical grouping variable with multiple groups

```{r}
anova_result <- aov(body_mass_g ~ species, data = penguin_data)
summary(anova_result)
```

### Correlation

Correlation measures the strength and direction of relationship between two numeric variables. Example: Does bill length increase with flipper length?

```{r}
cor(penguin_data$bill_length_mm,
    penguin_data$flipper_length_mm,
    use = "complete.obs" ) #removes missing values
```

### Linear Regression

Linear regression models the relationship between:

-    one predictor variable (X)

-    one outcome variable (Y)

It helps predict values and identify trends.

```{r}
model <- lm(flipper_length_mm ~ bill_length_mm, data = penguin_data)
summary(model)
```

### plot with regression line

```{r}
ggplot(penguin_data,
       aes(x = bill_length_mm,
           y = flipper_length_mm)) +
  geom_point() +
  geom_smooth(method = "lm")
```
